pdp: An R Package for Constructing Partial Dependence Plots

### 2016-12-11 manuscript/greenwell.tex
tweaks based on review comments

The \pkg{pdp} package is useful for constructing PDPs for many classes of fitted models in R. PDPs are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest. The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}:
\begin{example}
install.packages("pdp")
\end{example}
The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}. Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}. Currently, only three functions are exported by \pkg{pdp}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
  \item \code{multiplot}
\end{itemize}

@@

Finally, \code{multiplot} is a simple helper function that makes it easier to display multiple PDPs on a single plot.

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
multiplot(
  # Figure 5 (left)
  partial(boston.rf, "rm", plot = TRUE),

@@

# Train an XGBoost model
set.seed(202)
boston.xgb.tune <- train(x = data.matrix(subset(boston, select = -cmedv)),
                         y = boston$cmedv,
                         method = "xgbTree",
                         metric = "RMSE",
                         trControl = ctrl,
                         tuneGrid = xgb.grid)
\end{example}
The final model achieves a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
multiplot(
  partial(boston.xgb.tune, pred.var = "rm", plot = T, rug = T),
  partial(boston.xgb.tune, pred.var = "lstat", plot = T, rug = T),
  partial(boston.xgb.tune, pred.var = c("lstat", "rm"), plot = T, chull = T),
  ncol = 3)
\end{example}

>>>>

The \pkg{pdp} package is useful for constructing PDPs for many classes of fitted models in R. PDPs are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest. The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}:
\begin{example}
install.packages("pdp")
\end{example}
The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}. Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}. Currently, only two functions are exported by \pkg{pdp}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}

@@

\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}. \pkg{grid} graphics behave a little differently than traditional R graphics and two points are worth making:
\begin{enumerate}
  \item \pkg{lattice} functions create a \code{"trellis"} object, but does not display it; the \code{print} method produces the actual display. The result is automatically printed when using these functions in the command line, but in \code{source} or inside a function, an explicit \code{print} statement is required to display the resulting graph.
  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots. Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
\end{enumerate}
A consequence of second point is that the \code{par} function can not be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots. Simple solutions are available in packages \CRANpkg{latticeExtra} and \CRANpkg{gridExtra}. For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{pdp} plots on a single page.

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
grid.arrange(
  # Figure 5 (left)
  partial(boston.rf, "rm", plot = TRUE),

@@

# Train an XGBoost model
set.seed(202)
boston.xgb.tune <- train(x = data.matrix(subset(boston, select = -cmedv)),
                         y = boston$cmedv,
                         method = "xgbTree",
                         metric = "RMSE",
                         trControl = ctrl,
                         tuneGrid = xgb.grid)
\end{example}
The final model achieves a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
grid.arrange(
  # Figure 8 (left)
  partial(boston.xgb.tune, pred.var = "rm", plot = T, rug = T),

  # Figure 8 (middle)
  partial(boston.xgb.tune, pred.var = "lstat", plot = T, rug = T),

  # Figure 8 (right)
  partial(boston.xgb.tune, pred.var = c("lstat", "rm"), plot = T, chull = T),
